---
title: "High Cholesterol Risk Factors"
format: 
  dashboard:
    sidebar:
      width: 200px
    navbar:
      right:
        -text: "Blog"
        href: "blog/report1.html"
        # -text: "Home"
        # href: 
logo: images/logo.jpg
theme: [sandstone, theme/custom.scss]
fig-width: 10
fig-asp: 0.3
params:
  month: "March"
  year: "2020"
editor_options: 
  chunk_output_type: console
---

```{r}
#| label: load-packages
#| message: false

library(tidyverse)
library(readxl)
library(scales)
library(DT)
library(gt)
theme_set(theme_minimal(base_size = 24, base_family = "Atkinson Hyperlegible"))

source("Zhang431.R")
```

```{r}
#| label: load-data
#| message: false
# demo_raw <- read_xpt("P_DEMO.xpt")
# saveRDS(demo_raw, "data/P_DEMO.Rds")
# 
# wh_raw <- read_xpt("P_WHQ.xpt")
# saveRDS(wh_raw, "data/P_WHQ.Rds")
# 
# bp_raw <- read_xpt("P_BPQ.xpt")
# saveRDS(bp_raw, "data/P_BPQ.Rds")
#
# demo_raw <- readRDS("data/P_DEMO.Rds")
# wh_raw <- readRDS("data/P_WHQ.Rds")
# bp_raw <- readRDS("data/P_BPQ.Rds")
# 
# data_raw <- demo_raw %>%
#   left_join(wh_raw, by = "SEQN") %>%
#   left_join(bp_raw, by = "SEQN")
# 
# saveRDS(data_raw, "data/merged.Rds")
#
data_raw <- readRDS("data/merged.Rds")
```

```{r}
#| label: set-inputs

time_period <- paste(params$month, params$year)
```

```{r}
#| label: prep-data

data_clean <- data_raw %>%
  filter(RIDSTATR == 2) %>%
  filter(RIDAGEYR >= 21 & RIDAGEYR <= 79)

data_clean <- data_clean %>%
  select(SEQN, RIAGENDR, DMDEDUC2, WHD020, BPQ080) %>%
  mutate(
    # Recode DMDEDUC2 into four categories
    DMDEDUC2 = case_when(
      DMDEDUC2 %in% c(1, 2) ~ "Less than HS",
      DMDEDUC2 == 3 ~ "HS Grad",
      DMDEDUC2 == 4 ~ "Some College",
      DMDEDUC2 == 5 ~ "College Grad",
      TRUE ~ NA_character_
    ),
    # Handle missing values
    BPQ080 = ifelse(BPQ080 %in% c(7, 9), NA, BPQ080),
    WHD020 = ifelse(WHD020 %in% c(7777, 9999), NA, WHD020),
    # Convert to factor
    RIAGENDR = factor(RIAGENDR, levels = c(1, 2), labels = c("Male", "Female")),
    DMDEDUC2 = factor(DMDEDUC2, levels = c("Less than HS", "HS Grad", "Some College", "College Grad")),
    BPQ080 = factor(BPQ080, levels = c(1, 2), labels = c("Yes", "No"))
  )

data_clean <- data_clean[complete.cases(data_clean), ]
```

#  {.sidebar}


The data is obtained on `{r} time_period`.


::: {.callout-note collapse="true"}
## Data source

The data for this anlysis is sourced from  [The National Health and Nutrition Examination Survey (NHANES)](https://www.cdc.gov/nchs/nhanes/about_nhanes.htm).

NHANES is a program of studies designed to assess the health and nutritional status of adults and children in the United States. The survey is unique in that it combines interviews and physical examinations. NHANES is a major program of the National Center for Health Statistics (NCHS). NCHS is part of the Centers for Disease Control and Prevention (CDC) and has the responsibility for producing vital and health statistics for the Nation.

The NHANES interview includes demographic, socioeconomic, dietary, and health-related questions. The examination component consists of medical, dental, and physiological measurements, as well as laboratory tests administered by highly trained medical personnel.

For this study, several variables have been selected to explore potential risk factors associated with high cholesterol(listed below).
:::


| Variable   | Type    | 
|:----------:|:-------:|
| **ID**   | --      |
| **Gender**| Binary | 
| **Education level**| 4-cat  | 
| **Weight** | Quant   |
| **High cholesterol** | Binary  |

# Overview


```{r}
#| label: all-values
#| results: hide

n_total <- nrow(data_clean)

p_high <- data_clean |>
  count(BPQ080) |>
  mutate(p = n / sum(n)) |>
  filter(BPQ080 == "Yes") |>
  pull(p)

p_low <- data_clean |>
  count(BPQ080) |>
  mutate(p = n / sum(n)) |>
  filter(BPQ080 == "No") |>
  pull(p)
```


In this study, we aimed to explore the influencing factors of **High Cholesterol**. The candidate variables included:

- **Gender**: Male, Female.

- **Education Level**: Less than HS, HS Grad, Some College, College Grad.

- **Weight**: Quantitative measure in pounds.

## Row {height="40%"}

### Column {width="40%"}

```{r}
#| title: Numerical summary

data_clean |>
  count(BPQ080) |>
  mutate(p = n / sum(n)) |>
  gt() |>
  fmt_percent(
    columns = p,
    decimals = 1
  ) |>
  tab_style(
    style = cell_text(color = "#ae8b2d", weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = BPQ080 == "Yes"
    )
  ) |>
  tab_style(
    style = cell_text(color = "#0e2635", weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = BPQ080 == "No"
    )
  ) |>
  cols_label(
    BPQ080 = "",
    n = "Number of participants",
    p = "Proportion of high cholesterol",
    .fn = md
  )
```

### Column {width="60%"}

```{r}
#| title: Weight and high cholesterol

ggplot(data_clean, aes(x = BPQ080, y = WHD020, fill = BPQ080)) +
  geom_violin(alpha = 0.25) +
  geom_boxplot() +
  stat_summary(fun = "mean") +
  scale_fill_manual(
    values = c("#ae8b2d", "#0e2635"),
    guide = guide_legend(reverse = TRUE)
    ) +
  labs(x = "Cholesterol Status",
       y = "Weight (Pounds)",
       fill = "High Cholesterol") +
  theme_minimal()
```

## Row {height="35%"}

```{r}
#| title: Gender and high cholesterol

table_2x2 <- table(data_clean$RIAGENDR, 
                   data_clean$BPQ080, 
                   dnn = c("Gender", "High Cholesterol"))

table_count <- as.data.frame(table_2x2)
colnames(table_count) <- c("Gender", "High_Cholesterol", "Count")

ggplot(table_count, aes(x = Count, y = Gender, fill = High_Cholesterol)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(
    values = c("#ae8b2d", "#0e2635"),
    guide = guide_legend(reverse = TRUE)
    ) +
  scale_x_continuous(labels = label_percent()) +
  labs(y = NULL, x = NULL, fill = "High Cholesterol") +
  theme_minimal()
```

```{r}
#| title: Education level and high cholesterol

table_2x4 <- table(data_clean$BPQ080, 
                   data_clean$DMDEDUC2, 
                   dnn = c("High Cholesterol", "Education Level"))

table_count <- as.data.frame(table_2x4)
colnames(table_count) <- c("High_Cholesterol", "Education_Level", "Count")

ggplot(table_count, aes(x = Count, y = Education_Level, fill = High_Cholesterol)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(
    values = c("#ae8b2d", "#0e2635"),
    guide = guide_legend(reverse = TRUE)
    ) +
  scale_x_continuous(labels = label_percent()) +
  labs(y = NULL, x = NULL, fill = "High Cholesterol") +
  theme_minimal()
```

# Data

```{r}
data_clean |>
  arrange(SEQN) |>
  datatable(
    colnames = c("ID", "Gender", "Education level", "Weights (lb)", "High cholesterol"),
    options = list(dom = 'ftp', paging = TRUE)
    )
```

# Gender {orientation="columns"}

## Column {width="60%"}

### Row {height="35%"}

```{r}
#| title: Numerical summary

table_2x2 <- table(data_clean$BPQ080, 
                   data_clean$RIAGENDR, 
                   dnn = c("High Cholesterol", "Gender"))

formatted_table <- as.data.frame.matrix(table_2x2) %>%
  rownames_to_column(var = "High Cholesterol")

formatted_table %>%
  gt() %>%
  tab_spanner(
    label = "Gender",
    columns = c(Male, Female)
  ) %>%
  tab_style(
    style = cell_text(color = "#ae8b2d", weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = `High Cholesterol` == "Yes"
    )
  ) %>%
  tab_style(
    style = cell_text(color = "#0e2635", weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = `High Cholesterol` == "No"
    )
  ) %>%
  cols_label(
    `High Cholesterol` = "High Cholesterol",
    Male = "Male",
    Female = "Female"
  )

```

### Row {height="65%"}

```{r}
#| title: 2 by 2 table analysis

result <- twobytwo(
  a = 1370, b = 2344, c = 1326, d = 2623,
  namer1 = "Male", namer2 = "Female",
  namec1 = "High Cholesterol", namec2 = "Not High Cholesterol",
  conf.level = 0.95
)
```

```{r}
#| title: Maternal age and term

# ld_v |>
#   count(maternal_age, term) |>
#   ggplot(aes(x = n, y = fct_rev(maternal_age), fill = fct_rev(term))) +
#   geom_col(position = "fill", color = "white") +
#   scale_fill_brewer(
#     palette = "Greens",
#     guide = guide_legend(reverse = TRUE),
#     direction = -1
#   ) +
#   scale_x_continuous(labels = label_percent()) +
#   labs(y = NULL, x = NULL, fill = "Term")
```

## Column {width="40%"}

```{r}
#| title: Data

data_clean |>
  select(SEQN, RIAGENDR, BPQ080) |>
  arrange(SEQN) |>
  datatable(
    colnames = c("ID", "Gender", "High cholesterol"),
    options = list(
      dom = "lrtip",
      paging = TRUE,
      lengthMenu = c(5, 10, 25),
      pageLength = 10,
      scrollY = TRUE
    ),
  )
```

# Education {orientation="columns"}

## Column {width="60%"}

```{r}
#| label: cesarean-values
#| results: hide

# ld_c <- ld |>
#   filter(delivery_method == "Cesarean")
# 
# n_births_c <- nrow(ld_c)
# 
# p_preterm_c <- ld_c |>
#   count(term) |>
#   mutate(p = n / sum(n)) |>
#   filter(term == "Pre-term") |>
#   pull(p)
# 
# p_preterm_color_c <- case_when(
#   between(p_preterm_c, params$us_preterm_rate, params$us_preterm_rate + params$threshold_diff) ~ "warning",
#   p_preterm_c > params$us_preterm_rate + params$threshold_diff ~ "danger",
#   .default = "light"
#   )
```


### Row {height="50%"}

```{r}
#| title: Numerical summary

table_2x4 <- table(data_clean$BPQ080, 
                   data_clean$DMDEDUC2, 
                   dnn = c("High Cholesterol", "Education level"))

formatted_table_2x4 <- as.data.frame.matrix(table_2x4) %>%
  rownames_to_column(var = "High Cholesterol")

formatted_table_2x4 %>%
  gt(rowname_col = "High Cholesterol") %>%
  tab_spanner(
    label = "Education level",
    columns = everything()
  ) %>%
  tab_style(
    style = cell_text(color = "#ae8b2d", weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = `High Cholesterol` == "Yes"
    )
  ) %>%
  tab_style(
    style = cell_text(color = "#0e2635", weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = `High Cholesterol` == "No"
    )
  ) %>%
  cols_label(
    `High Cholesterol` = "High Cholesterol"
  )

```

### Row {height="50%"}

```{r}
#| title: Pearson's Chi-squared test

chi_test <- chisq.test(table_2x4)
chi_test
```



## Column {width="60%"}

```{r}
#| title: Data

data_clean |>
  select(SEQN, DMDEDUC2, BPQ080) |>
  arrange(SEQN) |>
  datatable(
    colnames = c("ID", "Education level", "High cholesterol"),
    options = list(
      dom = "lrtip",
      paging = TRUE,
      lengthMenu = c(5, 10, 25),
      pageLength = 10,
      scrollY = TRUE
    ),
  )
```

# Weight {orientation="columns"}

## Column {width="60%"}


```{r}
#| title: Weight distribution

ggplot(data_clean, aes(x = WHD020, fill = BPQ080)) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 20) +
  theme_minimal() +
  labs(x = "Weight (lb)") +
  scale_fill_manual(
    values = c("#ae8b2d", "#0e2635"),
    guide = guide_legend(reverse = TRUE)
    ) +
  xlim(50, 500)
```

```{r}
#| title: Numeric summary

summary_data <- data_clean %>%
  group_by(BPQ080) %>%
  summarise(
    mean = round(mean(WHD020),1),
    sd = round(sd(WHD020),1),
    median = median(WHD020),
    min = min(WHD020),
    q25 = round(quantile(WHD020, 0.25),1),
    q75 = round(quantile(WHD020, 0.75),1),
    max = max(WHD020)
  ) %>%
  rename(`High Cholesterol`= BPQ080)


summary_data %>%
  gt() %>%
  tab_spanner(
    label = "Weight Summary",
    columns = c(mean, sd, median, min, q25, q75, max)
  ) %>%
  tab_style(
    style = cell_text(color = "#ae8b2d", weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = `High Cholesterol` == "Yes"
    )
  ) %>%
  tab_style(
    style = cell_text(color = "#0e2635", weight = "bold"),
    locations = cells_body(
      columns = everything(),
      rows = `High Cholesterol` == "No"
    )
  )


```

## Column {width="40%"}

```{r}
#| title: Data

data_clean |>
  select(SEQN, WHD020, BPQ080) |>
  arrange(SEQN) |>
  datatable(
    colnames = c("ID", "Weight", "High cholesterol"),
    options = list(
      dom = "lrtip",
      paging = TRUE,
      lengthMenu = c(5, 10, 25),
      pageLength = 10,
      scrollY = TRUE
    ),
  )
```



